---
title: "MultiAssaySpatialExperiment Cheatsheet"
author: "Patrick Aboyoun"
date: "`r format(Sys.Date(), '%b %d, %Y')`"
output:
  BiocStyle::html_document:
    toc: true
    number_sections: true
    toc_depth: 2
    toc_float:
      collapsed: true
vignette: >
  %\VignetteIndexEntry{MultiAssaySpatialExperiment Cheatsheet}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Quick reference

MultiAssaySpatialExperiment (MASE) is a Bioconductor container for multi-assay
spatial omics data. It extends `MultiAssayExperiment` with spatial layers:
points, shapes, images, labels, and spatial mapping.

```{r load}
library(MultiAssaySpatialExperiment)
```

# Constructors

## Create MASE from scratch

```{r constructor}
# Minimal MASE with one assay
mat <- matrix(rnorm(20), nrow = 5, ncol = 4,
  dimnames = list(paste0("G", 1:5), paste0("S", 1:4)))

mase <- MultiAssaySpatialExperiment(
  experiments = ExperimentList(rna = mat),
  colData = DataFrame(row.names = paste0("S", 1:4)),
  sampleMap = DataFrame(
    assay = factor("rna"),
    primary = paste0("S", 1:4),
    colname = paste0("S", 1:4)
  )
)
```

## Add spatial layers

```{r add_spatial}
library(sf)

# Points layer (coordinates)
pts <- DataFrame(
  x = c(1, 2, 3, 4),
  y = c(1, 2, 3, 4),
  instance_id = paste0("S", 1:4)
)

# Shapes layer (polygons)
poly1 <- st_polygon(list(matrix(c(0, 0, 1, 0, 1, 1, 0, 1, 0, 0), ncol = 2, byrow = TRUE)))
shapes <- DataFrame(
  instance_id = "region1",
  geometry = st_sfc(poly1)
)

mase <- MultiAssaySpatialExperiment(
  experiments = ExperimentList(rna = mat),
  colData = DataFrame(row.names = paste0("S", 1:4)),
  sampleMap = DataFrame(
    assay = factor("rna"),
    primary = paste0("S", 1:4),
    colname = paste0("S", 1:4)
  ),
  points = PointsLayerList(coords = pts),
  shapes = ShapesLayerList(cells = shapes),
  spatialMap = DataFrame(
    assay = factor("rna"),
    colname = paste0("S", 1:4),
    element_type = "points",
    region = "coords",
    instance_id = paste0("S", 1:4)
  )
)
```

## Using prepMASE and buildSpatialMap

```{r constructor_prep}
sample_map <- DataFrame(
  assay = factor("rna"),
  primary = paste0("S", 1:4),
  colname = paste0("S", 1:4)
)
spmap <- buildSpatialMap(sample_map, region = "coords", element_type = "points")
prepared <- prepMASE(
  ExperimentList(rna = mat),
  DataFrame(row.names = paste0("S", 1:4)),
  sample_map,
  points = PointsLayerList(coords = pts),
  spatialMap = spmap
)
mase <- do.call(MultiAssaySpatialExperiment, prepared)
```

## Read from vendor outputs

```{r read_data}
# Read Xenium data
mase_xenium <- readXeniumMASE(
  data_dir = "path/to/xenium_output",
  sample_id = "sample1",
  segmentations = "cell",
  add_transcripts = FALSE
)

# Read Visium data
mase_visium <- readVisiumMASE(
  data_dir = "path/to/spaceranger_output",
  sample_id = "sample1",
  images = TRUE,
  data = "filtered"
)

# Read Visium HD data (separate reader; bin sizes "002", "008", "016")
mase_visium_hd <- readVisiumHDMASE(
  data_dir = "path/to/spaceranger_hd_output",
  sample_id = "sample_hd",
  bin_size = "008"
)

# Read CosMx data
mase_cosmx <- readCosMxMASE(
  data_dir = "path/to/cosmx_output",
  sample_id = "cosmx_sample",
  fov_ids = NULL  # NULL loads all FOVs
)

# Read Vizgen MERSCOPE data (MERFISH platform)
mase_merscope <- readMERSCOPEMASE(
  data_dir = "path/to/vizgen_output",
  sample_id = "merscope_sample",
  segmentation = "cellpose",
  load_transcripts = FALSE
)
```

# Accessors

## Inherited from MultiAssayExperiment

| Accessor | Returns |
|----------|---------|
| `experiments(mase)` | List of assays (ExperimentList) |
| `colData(mase)` | Specimen metadata (DataFrame) |
| `sampleMap(mase)` | Links specimens to assay columns (DataFrame) |
| `metadata(mase)` | Study-level metadata (list) |

## MASE-specific spatial accessors

| Accessor | Returns |
|----------|---------|
| `spatialPoints(mase)` | Points layers (PointsLayerList) |
| `spatialShapes(mase)` | Shapes layers (ShapesLayerList) |
| `spatialImages(mase)` | Images layers (RasterLayerList) |
| `spatialLabels(mase)` | Labels layers (RasterLayerList) |
| `spatialMap(mase)` | Links assay columns to spatial elements (DataFrame) |
| `imgData(mase)` | Image metadata (DataFrame) |

## Extract individual layers

```{r extract_layers}
# Get all points layers
pts_all <- spatialPoints(mase)

# Get specific layer by name
coords <- spatialPoints(mase)[["coords"]]

# Get shapes
cells <- spatialShapes(mase)[["cells"]]
```

## Setters

```{r setters}
# Replace points layer
spatialPoints(mase)[["coords"]] <- new_pts

# Replace entire points list
spatialPoints(mase) <- PointsLayerList(coords = pts, nuclei = nuclei_pts)

# Same pattern for shapes, images, labels, spatialMap, imgData
```

# Subsetting

MASE uses the same `subsetBy*` API as `MultiAssayExperiment`. There is no
separate spatial query function — spatial layers propagate automatically.

| Goal | Call | `y` type | Spatial propagation |
|------|------|----------|---------------------|
| Specimens | `subsetByColData(mase, y)` / `mase[, j, ]` | character, logical, numeric | `spatialMap`, `imgData`, points, shapes, images, labels |
| Observations | `subsetByColumn(mase, list(...))` | named list per assay | same as specimen path |
| Assays | `subsetByAssay(mase, y)` / `mase[, , k]` | character, logical, numeric | layers linked via `spatialMap`; drops recorded |
| Features | `subsetByRow(mase, y)` / `mase[i, , ]` | per-assay indices | spatial layers unchanged |
| Bounding box | `subsetByBoundingBox(mase, ...)` | coordinates | geometry filter → `subsetByColumn(list)` |
| Polygon | `subsetByPolygon(mase, polygon)` | sf geometry | geometry filter → `subsetByColumn(list)` |

## By row, column, assay

```{r subset_bracket}
# Bracket notation: mase[rows, cols, assays]
mase[1:3, , ]                  # First 3 features
mase[, c("S1", "S2"), ]        # Specimens S1, S2
mase[, , "rna"]                # Assay "rna"
mase[1:3, c("S1", "S2"), "rna"] # Combined
```

## By spatial region

```{r subset_spatial}
# Bounding box
mase_roi <- subsetByBoundingBox(mase, xmin = 0, xmax = 100, ymin = 0, ymax = 100)

# Polygon
library(sf)
poly <- st_polygon(list(matrix(c(0, 0, 100, 0, 100, 100, 0, 100, 0, 0), ncol = 2, byrow = TRUE)))
mase_poly <- subsetByPolygon(mase, poly)
```

## By assay

```{r subset_assay}
# Keep only specific assays (drops spatial layers not linked in spatialMap)
mase_rna <- subsetByAssay(mase, "rna")
```

## By colData (specimens)

```{r subset_coldata}
# Keep specific specimens
mase_subset <- subsetByColData(mase, c("S1", "S3"))
```

## By assay colData (observations)

```{r subset_column_metadata}
# Filter columns of one assay using experiment colData (MAE list interface)
cdf <- colData(experiments(mase)[["rna"]])
mase_sub <- subsetByColumn(mase, list(rna = cdf$region == "core"))
```

# Operations

## Combining

```{r combine}
# Combine two MASE objects (c method)
mase_combined <- c(mase1, mase2)
```

## Data extraction

```{r extract}
library(BiocGenerics)

# Long format (for plotting, analysis)
long_df <- longForm(mase)

# Wide format
wide_df <- wideFormat(mase)

# Complete cases (specimens with data in all assays)
complete_specimens <- complete.cases(mase)
```

## Intersecting

```{r intersect}
# Keep rows (features) present in all assays
mase_intersect_rows <- intersectRows(mase)

# Keep columns (specimens) present in all assays
mase_intersect_cols <- intersectColumns(mase)
```

## Replicates

```{r replicates}
# Merge technical replicates (list of replicate groups per assay)
mase_merged <- mergeReplicates(mase,
  replicates = list(rna = c("S1", "S1_rep")),
  simplify = BiocGenerics::mean)
```

# Spatial operations

## Annotate with regions

```{r annotate}
# Link points to shapes (spatial join)
mase <- annotateWithRegions(mase, points = "coords", shapes = "cells")

# spatialMap now has a "cells" column showing which cell each point is in
spatialMap(mase)
```

## Aggregate by region

```{r aggregate}
# Sum expression per cell
agg_sum <- aggregateByRegion(mase, by = "cells", FUN = "sum")

# Mean expression per cell
agg_mean <- aggregateByRegion(mase, by = "cells", FUN = "mean")

# Count points per cell
agg_count <- aggregateByRegion(mase, by = "cells", FUN = "count")
```

## Spatial join between layers

```{r spatial_join}
# Join two shape layers (requires geometry columns)
joined <- spatialJoin(spatialShapes(mase)[["cells"]], other_shapes,
                      join = st_intersects)
```

# Coercion

## To/from SpatialExperiment

```{r coerce_spe}
# MASE → SpatialExperiment (requires exactly one SPE-compatible assay)
spe <- as(mase, "SpatialExperiment")

# SpatialExperiment → MASE
mase_from_spe <- as(spe, "MultiAssaySpatialExperiment")
```

## Extract a single experiment

```{r extract_experiment}
# Access one assay directly (SummarizedExperiment, SCE, etc.)
sce <- experiments(mase)[["rna"]]
```

# Methods list

```{r methods_list}
# All methods for MASE
methods(class = "MultiAssaySpatialExperiment")

# Inherited MultiAssayExperiment methods
methods(class = "MultiAssayExperiment")
```

# Quick workflows

## Workflow 1: Build MASE from SummarizedExperiment

```{r workflow_se}
# Start with SummarizedExperiment
se <- SummarizedExperiment(assays = list(counts = mat))

# Build MASE manually (coercion is supported from SpatialExperiment)
mase <- MultiAssaySpatialExperiment(
  experiments = ExperimentList(rna = se),
  colData = DataFrame(row.names = paste0("S", 1:4)),
  sampleMap = DataFrame(
    assay = factor("rna"),
    primary = paste0("S", 1:4),
    colname = paste0("S", 1:4)
  )
)

# Add spatial coordinates
coords <- DataFrame(x = runif(4), y = runif(4), instance_id = paste0("S", 1:4))
spatialPoints(mase) <- PointsLayerList(coords = coords)
```

## Workflow 2: Spatial subset + aggregation

```{r workflow_spatial}
# 1. Subset to region of interest
mase_roi <- subsetByBoundingBox(mase, xmin = 0, xmax = 500, ymin = 0, ymax = 500)

# 2. Annotate points with cell boundaries
mase_roi <- annotateWithRegions(mase_roi, points = "coords", shapes = "cells")

# 3. Aggregate transcripts per cell
cell_expr <- aggregateByRegion(mase_roi, by = "cells", FUN = "sum")

# 4. Downstream analysis (clustering, DE, etc.)
# Use cell_expr as a regular matrix/SummarizedExperiment
```

## Workflow 3: Multi-assay spatial integration

```{r workflow_multiassay}
# MASE with RNA + protein assays
mase_multi <- MultiAssaySpatialExperiment(
  experiments = ExperimentList(rna = rna_mat, protein = protein_mat),
  colData = coldata,
  sampleMap = samplemap,
  points = PointsLayerList(coords = coords),
  spatialMap = spatialmap
)

# Both assays share the same spatial coordinates
# Analyze co-localization, correlation, etc.
rna_expr <- experiments(mase_multi)$rna
protein_expr <- experiments(mase_multi)$protein

# Spatial correlation per gene-protein pair
# (custom analysis code here)
```

# See also

| Topic | Vignette |
|-------|----------|
| Overview and anatomy | Introduction to MultiAssaySpatialExperiment |
| Construction, subsetting, annotation | Working with MultiAssaySpatialExperiment |
| Reader workflows and integration | MultiAssaySpatialExperiment Use Cases |
| Quick reference | MultiAssaySpatialExperiment Cheatsheet |

# Session info

```{r sessionInfo, eval = TRUE}
sessionInfo()
```
