---
title: "Coercion from SpatialExperiment, SpatialFeatureExperiment, and SpatialData"
author: "Patrick Aboyoun"
date: "`r format(Sys.Date(), '%b %d, %Y')`"
output:
  BiocStyle::html_document:
    toc: true
    number_sections: true
    toc_depth: 3
    toc_float:
      collapsed: true
vignette: >
  %\VignetteIndexEntry{Coercion from SpatialExperiment, SpatialFeatureExperiment, and SpatialData}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  cache = TRUE
)
```

```{r load-optional-first, include = FALSE}
## SpatialFeatureExperiment and SpatialData methods are registered only when
## those packages are already loaded when MultiAssaySpatialExperiment loads.
## Load them first so coercion to/from SFE and SpatialData is available.
if (requireNamespace("SpatialFeatureExperiment", quietly = TRUE))
  library(SpatialFeatureExperiment)
if (requireNamespace("SpatialData", quietly = TRUE))
  library(SpatialData)
```

# What coercion is doing

When you move data between object types, you are effectively running an ETL
step: extract from one representation, transform the structure, load into
another. The structure changes. Some information is preserved, some is reshaped,
and some may be dropped. Understanding that mapping helps you choose the right
format for your workflow and avoid surprises.

MASE supports coercion to and from three spatial representations:
SpatialExperiment, SpatialFeatureExperiment, and SpatialData. Each has a
different way of storing linkage between assays and spatial elements. The
coercion logic translates between those models. This vignette explains the
mappings in plain terms and what you gain or lose in each direction.

# The three representations

| Type | Linkage model | Best for |
|------|---------------|----------|
| **SpatialExperiment / SpatialFeatureExperiment** | Coordinates live inside the experiment (spatialCoords, colGeometries). Single assay. | One experiment, one technology, R-focused analysis |
| **SpatialData** | Each table carries its own metadata (region, instance_key) pointing to a spatial element. Flat layers. | On-disk, Python-interop, OME-NGFF / Zarr |
| **MASE** | Central mapping tables (sampleMap, spatialMap). Multiple assays share spatial elements. | Multi-assay, subsetting with referential integrity, Parquet export |

# SpatialExperiment ↔ MASE

SpatialExperiment (and its subclass SpatialFeatureExperiment) stores spatial
coordinates inside the experiment: `spatialCoords` is a matrix with one row per
column, and `imgData` links samples to images. There is no separate "points" or
"shapes" table—the coordinates are part of the assay container.

## SPE → MASE: wrapping a single experiment

**What gets mapped:**

- The SPE becomes the sole assay in MASE's `ExperimentList`.
- `spatialCoords` (the matrix of x, y per column) is copied into a
  `PointsLayerList` entry named "coordinates". Each row gets an `instance_id`
  matching the assay column name.
- `imgData` is passed through to MASE's `imgData` slot unchanged.
- `sample_id` from colData (or `"sample01"` if absent) becomes the primary
  identifier in `sampleMap`. Each assay column is linked to its specimen.

**What you gain:** The multi-assay container. Even with one assay, you now have
`sampleMap`, shared `points` and `imgData` slots, and the option to add more
assays later. Subsetting by specimen will propagate correctly.

**What you lose:** Nothing essential. The coordinates move from `spatialCoords`
(inside the experiment) to the `points` slot (shared container). The data is the
same; the packaging changes. Note: the SPE→MASE coercion does not populate
`spatialMap` by default—the link between assay columns and the points element is
implicit (matching row order and colnames). If you need explicit `spatialMap`
rows for validity or for sharing points across assays, add them manually.

## MASE → SPE: unwrapping to a single experiment

**What gets mapped:**

- The single assay is extracted from `ExperimentList`. It must be (or coerce to)
  a SpatialExperiment or SpatialFeatureExperiment.
- MASE's `imgData`, if present, overwrites the assay's `imgData`.
- The assay is returned as-is. Its `spatialCoords` (or SFE geometries) stay
  where they are—they were put there when the assay was created or when it was
  wrapped from SPE.

**What you gain:** A standard SPE or SFE for single-assay pipelines (e.g.,
packages that expect one experiment).

**What you lose:** All other assays. If MASE had RNA, protein, and ATAC, you
keep only the one you extract. MASE's `spatialMap`, `points`, and `shapes` slots
are not written back into the experiment—the assay already carries its own
coordinates. If you built a MASE from scratch with points in the shared slot but
the assay never had `spatialCoords`, coercing to SPE would give you an SPE whose
spatial coordinates might be empty or derived from the assay's structure, not
from MASE's points. In practice, coercion to SPE works best when the MASE was
created by wrapping an SPE (or SFE) in the first place.

# SpatialFeatureExperiment ↔ MASE

SpatialFeatureExperiment extends SpatialExperiment with `sf`-based geometries
(colGeometries, rowGeometries, annotGeometries), spatial graphs, and local
spatial statistics. The coercion path is the same as for SPE: SFE is a subclass
of SPE, so SFE→MASE uses the SPE coercion (wrap the experiment, copy
`spatialCoords` to points). MASE→SFE extracts the single assay and requires it
to be an SFE or SPE (which is then coerced to SFE).

**What you gain / lose:** Same as SPE, with one caveat. SFE's richer geometries
(cell boundaries, tissue outlines, spatial graphs) live inside the experiment,
not in MASE's shared `points` or `shapes` slots. When you wrap an SFE in MASE,
only the basic `spatialCoords` are copied to `points`. The colGeometries,
annotGeometries, and spatial graphs stay in the assay. When you coerce back to
SFE, you get the full experiment back. So you do not lose SFE-specific data; it
just stays in the experiment layer rather than moving to MASE's shared geometry
slots.

# SpatialData ↔ MASE

SpatialData uses a different design. Tables (SingleCellExperiments) and spatial
elements (points, shapes, images, labels) are separate. Each table carries
`int_metadata$spatialdata_attrs` with `region` (which element this table
annotates) and `instance_key` (which colData column holds the instance id). The
linkage is per-table. MASE, by contrast, has a central `spatialMap` that links
(assay, colname) to (region, instance_id). Coercion is where we translate
between those two models.

## SpatialData → MASE: normalizing the linkage

**What gets mapped:**

- Each table in `tables(sd)` becomes one assay in MASE's `ExperimentList`. Table
  names become assay names.
- Points and shapes are materialized from PointFrame and ShapeFrame into
  DataFrames (arrow datasets are collected into memory).
- The per-table `spatialdata_attrs` (region, region_key, instance_key) are read
  from each SCE's `int_metadata`. For tables that annotate a points or shapes
  element, we build rows for the central `spatialMap`: (assay, colname, region,
  instance_id).
- The `sampleMap` is built so that each table's columns map to a primary equal
  to the table name (one primary per table). `colData` has one row per table
  name.

**What you gain:** Central mapping tables. No more scanning each table's
metadata to find linkage. Referential integrity checks. Subsetting and merging
that keep maps in sync. The in-memory MAE/MASE structure.

**What you lose:** Images and labels are not yet transferred—the SpatialData →
MASE path leaves those slots empty (ZarrArray handling is future work). If
points or shapes were on disk (Arrow), they are loaded into memory; large
datasets may use more RAM. The coordinate transform graph from SpatialData's
OME-NGFF layout is not yet carried over into MASE metadata.

## MASE → SpatialData: denormalizing for interoperability

**What gets mapped:**

- Each assay in MASE becomes one table. Assay names become table names.
- Points and shapes become PointFrame and ShapeFrame elements (in-memory
  `data.frame`s). For disk-backed SpatialData, you would write these to
  Arrow/Parquet separately.
- For each assay that has rows in `spatialMap`, we add
  `int_metadata$spatialdata_attrs` to the SCE: `region`, `region_key`,
  `instance_key`. We also add `region_key` and `instance_key` columns to colData
  if missing. The central map is "denormalized" back into per-table metadata;
  exactly what SpatialData expects.
- Images and labels are not transferred (would require ImageArray/LabelArray
  conversion).

**What you gain:** A SpatialData object for Python spatialdata compatibility,
OME-NGFF tooling, and workflows that expect the flat, per-table linkage model.

**What you lose:** Images and labels. The central `spatialMap` is distributed
into each table's metadata—you no longer have one place to query linkage. The
PointFrame and ShapeFrame elements are in-memory; you do not get automatic
Arrow/Parquet writing.

# Round-tripping

**SPE → MASE → SPE:** Preserves the assay, spatialCoords (via points), and
imgData. Works as expected when you started from an SPE.

**SpatialData → MASE → SpatialData:** Tables, points, and shapes round-trip. The
per-table `spatialdata_attrs` is reconstructed from `spatialMap` on the return
leg. Images and labels do not round-trip (both directions drop them for now).

# Summary table

| Direction | Preserved | Gained | Lost or changed |
|-----------|-----------|--------|-----------------|
| SPE → MASE | Assay, coords (→ points), imgData | sampleMap, points slot, multi-assay container | — |
| MASE → SPE | Single assay (with its own coords), imgData | SPE structure for single-assay tools | Other assays, explicit spatialMap not written back |
| SD → MASE | Tables, points, shapes (materialized) | Central spatialMap, referential integrity | Images, labels, on-disk/lazy evaluation |
| MASE → SD | Assays (→ tables), points, shapes | SpatialData structure, Python interop | Images, labels; central map becomes per-table metadata |

# SpatialExperiment example

```{r spe-to-mase}
library(MultiAssaySpatialExperiment)
library(SpatialExperiment)

mat <- matrix(1:12, 3, 4)
colnames(mat) <- paste0("cell", 1:4)
spe <- SpatialExperiment(
  assays = list(counts = mat),
  spatialCoords = matrix(c(1:4, 5:8), 4, 2)
)

mase <- as(spe, "MultiAssaySpatialExperiment")
mase
spatialPoints(mase)
```

## MASE → SPE

```{r mase-to-spe}
spe2 <- as(mase, "SpatialExperiment")
spe2
validObject(spe2)
```

# SpatialFeatureExperiment example

When SpatialFeatureExperiment is installed and loaded before MASE, SFE coerces to
MASE via the SPE path (SFE extends SPE). Reverse coercion extracts the single
assay and promotes it to SFE if needed. The vignette loads SFE in the setup so
that MASE registers the coercion methods at load time.

```{r sfe-coercion, eval = requireNamespace("SpatialFeatureExperiment", quietly = TRUE)}
sfe <- SpatialFeatureExperiment(
  assays = list(counts = matrix(1:8, 2, 4)),
  spatialCoords = matrix(rnorm(8), 4, 2),
  colData = S4Vectors::DataFrame(row.names = paste0("cell", 1:4))
)
mase_sfe <- as(sfe, "MultiAssaySpatialExperiment")
sfe2 <- as(mase_sfe, "SpatialFeatureExperiment")
validObject(sfe2)
```

# SpatialData example

When SpatialData is installed and loaded before MASE, coercion to/from
SpatialData is available. The vignette loads SpatialData in the setup so that
MASE registers the coercion methods at load time.

## SpatialData → MASE

```{r sd-to-mase, eval = requireNamespace("SpatialData", quietly = TRUE)}
library(SingleCellExperiment)
sce <- SingleCellExperiment(
  assays = list(counts = matrix(1:6, 2, 3)),
  colData = S4Vectors::DataFrame(row.names = paste0("cell", 1:3))
)
sd <- SpatialData::SpatialData(
  images = list(), labels = list(), points = list(), shapes = list(),
  tables = list(assay1 = sce)
)
mase <- as(sd, "MultiAssaySpatialExperiment")
mase
```

This minimal SpatialData has no points or shapes, and no `spatialdata_attrs` on
the table, so `spatialMap` will be NULL. To see the full SD→MASE mapping with
`spatialMap`, use a SpatialData object whose tables have
`int_metadata$spatialdata_attrs` pointing to points or shapes elements.

## MASE → SpatialData

```{r mase-to-sd, eval = requireNamespace("SpatialData", quietly = TRUE)}
sd2 <- as(mase, "SpatialData")
SpatialData::tableNames(sd2)
```

# Session info

```{r sessionInfo}
sessionInfo()
```
