---
title: "Introduction to MultiAssaySpatialExperiment"
author: "Patrick Aboyoun"
date: "`r format(Sys.Date(), '%b %d, %Y')`"
output:
  BiocStyle::html_document:
    toc: true
    number_sections: true
    toc_depth: 3
    toc_float:
      collapsed: true
vignette: >
  %\VignetteIndexEntry{Introduction to MultiAssaySpatialExperiment}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  cache = TRUE
)
```

# Why another spatial data structure?

If you have spent a few decades designing databases, you have seen the same
story many times: someone builds a system that works for one use case, and it
spreads. Later, a new use case appears that does not quite fit. People bolt on
workarounds, and the design gets messy. Eventually, someone steps back and asks:
what would we build if we started from the problem we actually have?

Spatial omics is at that stage. We already have excellent tools for
*single-assay* spatial data: SpatialExperiment and SpatialFeatureExperiment
handle one experiment at a time—one technology, one matrix, one set of spatial
coordinates. They work well when your question is "what genes are expressed in
this Visium section?" or "where are these MERFISH cells in tissue?" But
increasingly, people ask questions that cut across assays: "I have RNA, protein,
and chromatin from the same tissue—how do I link them and their spatial
coordinates in one place?" That is a different problem. It needs a *multi-assay*
structure that can carry spatial context and keep everything consistent when you
subset, merge, or query.

MultiAssayExperiment (MAE) solves the multi-assay part. It gives you a central
mapping table—the `sampleMap`—that links assay columns to specimens. One source
of truth, explicit foreign keys, and validity checks that keep the links
correct. When you subset by specimen, the maps update. No orphaned rows. No
hunting through metadata to figure out which assay column belongs to which
sample.

What MAE does not have is spatial geometry: coordinates for cells or spots, cell
boundaries, tissue images. That is where MultiAssaySpatialExperiment (MASE)
comes in. MASE extends MAE with spatial elements—points, shapes, images,
labels—and a second central mapping table, the `spatialMap`, that links assay
columns to those elements. Same philosophy as MAE: one place to look, one place
to update, referential integrity enforced. When you subset by specimen or assay,
the spatial maps and element lists stay in sync. You do not have to remember to
clean up coordinates or drop orphaned geometries manually.

The alternative—used by SpatialData and the scverse ecosystem—stores the linkage
inside each table. Each assay carries its own metadata saying "I annotate region
X, use instance_key Y." To find out which assays have spatial data, you loop
over tables and peek at their metadata. To subset consistently, you must touch
every table and hope you did not miss one. It works, but it is not the way you
would design a database. MASE follows the Bioconductor approach: central mapping
tables, explicit foreign keys, and validity checks that guarantee the links stay
correct. If that sounds like good schema design to you, MASE is built for you.

# When to use MASE

Use MultiAssaySpatialExperiment when:

- You have **multiple spatial assays** from the same or related tissue (e.g.,
  transcriptomics, proteomics, ATAC-seq).
- You need **one object** that holds the experiments, the specimen metadata, the
  spatial elements (coordinates, shapes, images), and the links between them.
- You want **subsetting and merging** that keep all links consistent—no manual
  cleanup, no orphaned geometry.
- You need to **convert** to or from SpatialExperiment,
  SpatialFeatureExperiment, or SpatialData for downstream tools.

If you have a single assay with spatial data, SpatialExperiment or
SpatialFeatureExperiment may be enough. If you need on-disk,
Python-interoperable storage, SpatialData has that niche. MASE sits in the
middle: multi-assay, in-memory, and built on the
same relational principles as MAE.

# The structure: what MASE adds to MAE

MASE inherits the core of MultiAssayExperiment: `ExperimentList` (your assays),
`colData` (specimen metadata), and `sampleMap` (assay column → specimen). On top
of that, it adds:

| Slot | What it holds | How it links |
|------|---------------|--------------|
| **points** | Named sets of (x, y) coordinates | `spatialMap` links (assay, colname) → (region, instance_id) |
| **shapes** | Named sets of polygons or lines | Same as points, via `spatialMap` |
| **images** | Raster images (tissue photos, etc.) | Referenced by `imgData`, which links sample_id → image |
| **labels** | Segmentation masks (pixel = instance id) | Via `spatialMap` when linking masks to assay rows |
| **imgData** | Which image goes with which specimen | sample_id → `colData` rownames |
| **spatialMap** | Assay column → spatial element row | (assay, colname) → (region, instance_id) |

The experiments in `ExperimentList` can be SummarizedExperiment,
SingleCellExperiment, SpatialExperiment, or SpatialFeatureExperiment. The
spatial elements are shared across assays: multiple assays can reference the
same points or shapes. The `spatialMap` is the central table that says which
assay column maps to which row in which spatial element. One table, one source
of truth. When you subset, it updates. Validity checks enforce that every row in
`spatialMap` points to a valid `sampleMap` row and a valid row in the
corresponding points or shapes element.

# Quick start

```{r quickstart}
library(MultiAssaySpatialExperiment)
```

Construct a minimal MASE with one experiment and point coordinates:

```{r construct}
mat <- matrix(rnorm(20), nrow = 5, ncol = 4,
  dimnames = list(paste0("G", 1:5), paste0("S", 1:4)))
pts <- S4Vectors::DataFrame(
  x = c(1, 2, 3, 4),
  y = c(1, 2, 3, 4),
  instance_id = paste0("S", 1:4))

mase <- MultiAssaySpatialExperiment(
  experiments = ExperimentList(assay1 = mat),
  colData = S4Vectors::DataFrame(row.names = paste0("S", 1:4)),
  sampleMap = S4Vectors::DataFrame(
    assay = factor("assay1", "assay1"),
    primary = paste0("S", 1:4),
    colname = paste0("S", 1:4)
  ),
  points = PointsLayerList(coords = pts)
)

mase
```

Here, `sampleMap` says each assay column (S1, S2, S3, S4) maps to the specimen
with that name. The `points` element "coords" has one row per column, with
`instance_id` matching the colnames. When you add a `spatialMap`, it makes that
link explicit and allows multiple assays to share the same spatial element. For
this minimal example, the coordinates live in the points slot; a full
`spatialMap` would add the formal (assay, colname, region, instance_id) rows.

# Accessing spatial elements

Use `spatialPoints`, `spatialShapes`, `spatialImages`, and `spatialLabels` to
get the element lists, and the singular forms (e.g., `spatialPoint`) to get one
element by name:

```{r accessors}
spatialPoints(mase)
spatialPointNames(mase)
spatialPoint(mase, "coords")
```

# Coercion

MASE can convert to and from SpatialExperiment, SpatialFeatureExperiment, and
SpatialData. When coercing from SpatialData, the per-table metadata (region,
instance_key) is translated into the central `spatialMap`. When coercing back,
`spatialMap` is used to reconstruct that metadata. See the *Coercion* vignette
for details.

# Spatial annotation and aggregation

To annotate measurement points with the regions (cells, segments) they fall in,
and to aggregate assay values by region, use `annotateWithRegions` and
`aggregateByRegion`. See the *Spatial annotation and aggregation* vignette.

# Session info

```{r sessionInfo}
sessionInfo()
```
