---
title: "Spatial annotation and aggregation"
author: "Patrick Aboyoun"
date: "`r format(Sys.Date(), '%b %d, %Y')`"
output:
  BiocStyle::html_document:
    toc: true
    number_sections: true
    toc_depth: 3
    toc_float:
      collapsed: true
vignette: >
  %\VignetteIndexEntry{Spatial annotation and aggregation}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  cache = TRUE
)
```

# Overview

Spatial omics often measures gene expression or other signals at discrete
locations—spots, cell centroids, or nuclei—that lie within larger regions such
as cells, tissue segments, or anatomical structures. A common workflow is to:

1. **Annotate** each measurement point with the region (e.g., cell) it falls in.
2. **Aggregate** assay values by region—summing counts per cell, averaging
   expression per segment, or counting points per region.

MultiAssaySpatialExperiment provides `annotateWithRegions` for step 1 and
`aggregateByRegion` for step 2. Both use the central `spatialMap` table to link
assay columns to spatial elements (points and shapes).

# Prerequisites

You need points (coordinates of measurements) and shapes (polygons defining
regions). The points layer must have `instance_id` values that match
`spatialMap`; the shapes layer must have `instance_id` and a `geometry`
column. Run `annotateWithRegions` first—it adds a column to `spatialMap`
linking each point to the shape that contains it. Then `aggregateByRegion`
uses that column to group and aggregate assay values.

# Example setup

We construct a minimal MASE: four spots (transcriptomic measurements) with
coordinates, two cell polygons, and a gene expression assay. The goal is to
annotate each spot with its containing cell and then aggregate expression per
cell.

```{r setup-example}
library(MultiAssaySpatialExperiment)
library(sf)
```

## Points and shapes

Four spots at coordinates (1.5, 1.5), (2.5, 1.5), (2.5, 2.5), and (3.5, 2.5).
Their `instance_id` values (S1–S4) will link to assay columns.

Two cell polygons:
- **cell1**: axis-aligned box from (1, 1) to (2, 2)—contains S1.
- **cell2**: box from (2, 1) to (3, 2)—contains S2.
- **cell3**: box from (2, 2) to (3, 3)—contains S3.
- S4 at (3.5, 2.5) falls outside all cells.

```{r create-geometry}
pts <- S4Vectors::DataFrame(
  x = c(1.5, 2.5, 2.5, 3.5),
  y = c(1.5, 1.5, 2.5, 2.5),
  instance_id = paste0("S", 1:4))

cell1 <- sf::st_polygon(list(matrix(c(1, 1, 2, 1, 2, 2, 1, 2, 1, 1), ncol = 2, byrow = TRUE)))
cell2 <- sf::st_polygon(list(matrix(c(2, 1, 3, 1, 3, 2, 2, 2, 2, 1), ncol = 2, byrow = TRUE)))
cell3 <- sf::st_polygon(list(matrix(c(2, 2, 3, 2, 3, 3, 2, 3, 2, 2), ncol = 2, byrow = TRUE)))

shp_df <- S4Vectors::DataFrame(
  instance_id = c("cell1", "cell2", "cell3"),
  geometry = sf::st_sfc(cell1, cell2, cell3))
```

## Assay and spatialMap

A 3-gene × 4-spot expression matrix. The `spatialMap` links each assay column
(S1–S4) to the points layer "centroids" and its `instance_id`.

```{r create-mase}
expr <- matrix(c(10, 20, 5,  15,
                 30, 10, 25, 5,
                 5,  15, 20, 30),
  nrow = 3, ncol = 4,
  dimnames = list(paste0("Gene", 1:3), paste0("S", 1:4)))

mase <- MultiAssaySpatialExperiment(
  experiments = ExperimentList(rna = expr),
  colData = S4Vectors::DataFrame(row.names = paste0("S", 1:4)),
  sampleMap = S4Vectors::DataFrame(
    assay = factor("rna"),
    primary = paste0("S", 1:4),
    colname = paste0("S", 1:4)
  ),
  points = PointsLayerList(centroids = pts),
  shapes = ShapesLayerList(cells = shp_df),
  spatialMap = S4Vectors::DataFrame(
    assay = factor("rna"),
    colname = paste0("S", 1:4),
    region = "centroids",
    instance_id = paste0("S", 1:4)
  )
)
mase
```

# annotateWithRegions: spatial join

`annotateWithRegions` performs a spatial join: for each point, it finds which
shape (if any) matches and records the shape's `instance_id`. That annotation
is written to a new column in `spatialMap`, using the shapes layer name by
default. The `join` argument follows the convention of
[`sf::st_join`](https://r-spatial.github.io/sf/reference/st_join.html).

By default, `join = st_intersects` (point-in-polygon): points inside a shape
get that shape's ID; points outside all shapes get `NA`. Use
`join = sf::st_nearest_feature` to assign each point to its nearest shape when
no shape contains it.

```{r annotate}
mase_orig <- mase
mase <- annotateWithRegions(mase, points = "centroids", shapes = "cells")
spatialMap(mase)
```

The new column `cells` shows: S1 → cell1, S2 → cell2, S3 → cell3, S4 → NA
(no containing cell). To assign S4 to its nearest cell instead:

```{r annotate-nearest}
mase_nn <- annotateWithRegions(mase_orig, points = "centroids", shapes = "cells",
  join = sf::st_nearest_feature)
spatialMap(mase_nn)
```

You can pass `regionCol` to choose a different column name.

# aggregateByRegion: aggregate by shape

With the annotation in place, `aggregateByRegion` groups assay columns by
region and applies an aggregation function.

## Count points per region

`FUN = "count"` returns the number of points in each shape:

```{r aggregate-count}
agg_count <- aggregateByRegion(mase, by = "cells", FUN = "count")
agg_count
```

## Sum expression per region

`FUN = "sum"` sums each gene's expression over spots that fall in the same
cell. The result is a list of matrices (one per assay), with features as rows
and region IDs as columns:

```{r aggregate-sum}
agg_sum <- aggregateByRegion(mase, by = "cells", FUN = "sum")
agg_sum[["rna"]]
```

Interpretation: cell1 has one spot (S1), so the values are unchanged.
cell2 has S2; cell3 has S3. Rows with `NA` in the `by` column (S4) are excluded.

## Mean expression per region

`FUN = "mean"` averages expression across spots in each cell. With one spot per
cell, mean equals the original value; with multiple spots per cell, you get
the mean per gene per cell:

```{r aggregate-mean}
agg_mean <- aggregateByRegion(mase, by = "cells", FUN = "mean")
agg_mean[["rna"]]
```

# Multi-assay aggregation

When you have multiple assays (e.g., RNA and protein from the same spots),
`aggregateByRegion` aggregates all of them. Specify `assays` to limit which
assays are processed:

```{r multi-assay}
protein <- matrix(1:12, nrow = 2, ncol = 4,
  dimnames = list(paste0("P", 1:2), paste0("S", 1:4)))

mase2 <- MultiAssaySpatialExperiment(
  experiments = ExperimentList(rna = expr, protein = protein),
  colData = S4Vectors::DataFrame(row.names = paste0("S", 1:4)),
  sampleMap = S4Vectors::DataFrame(
    assay = factor(c(rep("rna", 4), rep("protein", 4))),
    primary = rep(paste0("S", 1:4), 2),
    colname = rep(paste0("S", 1:4), 2)
  ),
  points = PointsLayerList(centroids = pts),
  shapes = ShapesLayerList(cells = shp_df),
  spatialMap = S4Vectors::DataFrame(
    assay = factor(c(rep("rna", 4), rep("protein", 4))),
    colname = rep(paste0("S", 1:4), 2),
    region = "centroids",
    instance_id = rep(paste0("S", 1:4), 2)
  )
)
```

Both assays share the same spots (S1–S4) and `instance_id` in the points
layer. After `annotateWithRegions`, you can aggregate both assays:

```{r multi-agg}
mase2 <- annotateWithRegions(mase2, points = "centroids", shapes = "cells")
agg_both <- aggregateByRegion(mase2, by = "cells", FUN = "sum")
names(agg_both)
agg_both[["rna"]]
agg_both[["protein"]]
```

# Custom coordinate columns

If your points layer uses different column names for x and y (e.g., `X_centroid`,
`Y_centroid`), pass `spatialCoordsNames`:

```{r custom-coords}
pts2 <- S4Vectors::DataFrame(
  X_centroid = c(1.5, 2.5),
  Y_centroid = c(1.5, 2.5),
  instance_id = c("S1", "S2"))
# annotateWithRegions(mase, points = "centroids", shapes = "cells",
#   spatialCoordsNames = c("X_centroid", "Y_centroid"))
```

# See also

- **Subset vignette**: `subsetByBoundingBox` and `subsetByPolygon` use similar
  spatial logic and propagate to assays via `spatialMap`.
- **Introduction vignette**: MASE structure and the role of `spatialMap`.

# Session info

```{r sessionInfo}
sessionInfo()
```
