---
title: "Subsetting MultiAssaySpatialExperiment"
author: "Patrick Aboyoun"
date: "`r format(Sys.Date(), '%b %d, %Y')`"
output:
  BiocStyle::html_document:
    toc: true
    number_sections: true
    toc_depth: 3
    toc_float:
      collapsed: true
vignette: >
  %\VignetteIndexEntry{Subsetting MultiAssaySpatialExperiment}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  cache = TRUE
)
```

# Overview

MultiAssaySpatialExperiment (MASE) extends MultiAssayExperiment with spatial
slots—points, shapes, images, labels, `imgData`, and `spatialMap`. When you
subset a MASE, these slots must stay consistent with the assays and specimen
metadata. MASE's subsetting methods handle that propagation automatically.

This vignette describes each subsetting operation, shows examples, and provides
an **advanced section for developers** who want to support alternative data
representations (e.g., DuckDBDataFrame) in MASE's points and shapes slots.

# Subsetting operations

| Operation | What it filters | Spatial propagation |
|-----------|-----------------|---------------------|
| `subsetByColData(x, y)` | Specimens (primary identifiers) | `spatialMap`, `imgData` |
| `subsetByRow(x, y, ...)` | Rows of experiments | Spatial layers unchanged (assay-level) |
| `subsetByColumn(x, y)` | Assay columns | `spatialMap`, `imgData` |
| `subsetByAssay(x, y)` | Assays | Images, labels, points, shapes, `spatialMap`, `imgData` |
| `x[i, j, k, ..., drop]` | Row, column, assay indices | Composes the above |
| `subsetByBoundingBox(x, xmin, xmax, ymin, ymax, ...)` | Points/shapes by rectangle | Assays via `spatialMap` |
| `subsetByPolygon(x, polygon, ...)` | Points/shapes by polygon | Assays via `spatialMap` |

# Example: basic subsetting

```{r basic-setup}
library(MultiAssaySpatialExperiment)

mat <- matrix(rnorm(25), nrow = 5, ncol = 5,
  dimnames = list(paste0("G", 1:5), paste0("S", 1:5)))
pts <- S4Vectors::DataFrame(
  x = c(1, 2, 3, 4, 5),
  y = c(1, 2, 3, 4, 5),
  instance_id = paste0("S", 1:5))

mase <- MultiAssaySpatialExperiment(
  experiments = ExperimentList(assay1 = mat),
  colData = S4Vectors::DataFrame(row.names = paste0("S", 1:5)),
  sampleMap = S4Vectors::DataFrame(
    assay = factor("assay1"),
    primary = paste0("S", 1:5),
    colname = paste0("S", 1:5)
  ),
  points = PointsLayerList(coords = pts),
  spatialMap = S4Vectors::DataFrame(
    assay = factor("assay1"),
    colname = paste0("S", 1:5),
    region = "coords",
    instance_id = paste0("S", 1:5)
  )
)
mase
```

## subsetByColData — filter by specimen

Subset by primary identifiers (rownames of `colData`). Rows in `spatialMap` and
`imgData` that reference dropped specimens are removed.

```{r subsetByColData}
mase[, c("S1", "S3", "S5")]
```

## subsetByColumn — filter assay columns

Subset columns within each assay. When `y` is a list, each element applies to
the corresponding assay (named or by position). `spatialMap` and `imgData` are
filtered to retain only columns that remain.

```{r subsetByColumn}
mase[, list(assay1 = c("S2", "S4"))]
```

## subsetByAssay — filter assays

Keep only selected assays. Images, labels, points, shapes, `spatialMap`, and
`imgData` are trimmed to the retained assays. Points and shapes are filtered
by region name (only regions referenced in `spatialMap` after the assay filter
are kept).

```{r subsetByAssay}
mase[, , "assay1"]
```

## Bracket notation

The `[i, j, k, drop]` operator composes the above:

- `i` → rows (via `subsetByRow`)
- `j` → columns; vector → `subsetByColData`, list → `subsetByColumn`
- `k` → assays (via `subsetByAssay`)
- `drop = TRUE` → remove empty assays

```{r bracket}
mase[1:3, c("S1", "S2"), "assay1", drop = FALSE]
```

# Spatial subsetting

When you have `spatialMap` linking assay columns to spatial elements, you can
subset by **bounding box** or **polygon**. Points and shapes within the region
are kept; assays are filtered to retained instance_ids via `spatialMap`.

## subsetByBoundingBox

Filter by a rectangle `(xmin, xmax, ymin, ymax)`. Points need `x` and `y`
columns (customizable via `x_col`, `y_col`). Shapes need a `geometry` column
(sf `sfc`).

```{r subsetByBoundingBox}
m2 <- subsetByBoundingBox(mase, xmin = 1.5, xmax = 4.5, ymin = 1.5, ymax = 4.5)
m2
spatialPoints(m2)[["coords"]]
ncol(experiments(m2)[["assay1"]])
```

## subsetByPolygon

Filter by an sf polygon (`sfg`, `sfc`, or `sf`). Points and shapes intersecting
the polygon are retained.

```{r subsetByPolygon}
library(sf)
poly <- st_polygon(list(matrix(
  c(1.5, 1.5, 4.5, 1.5, 4.5, 4.5, 1.5, 4.5, 1.5, 1.5),
  ncol = 2, byrow = TRUE)))
m3 <- subsetByPolygon(mase, poly)
nrow(spatialPoints(m3)[["coords"]])
ncol(experiments(m3)[["assay1"]])
```

# Advanced: implementation guide for developers

MASE's subsetting uses only S3/S4 generics (verified via BiocGenerics and base R).
Table-like elements (`spatialMap`, `sampleMap`, `imgData`, points, shapes) and
layer lists are operated on through these generics. This section documents
which generics are used for each operation.

## Overview: generics by operation

All generics below are S3 or S4: `nrow`, `colnames`, `rownames`, `dim`, `names`, `length`, `unique` from BiocGenerics; `split` (S3); `[[`, `[` (primitives); `st_*` from sf (S3).

| Generic | subsetByColData | subsetByColumn | subsetByAssay | subsetByBbox | subsetByPolygon |
|---------|-----------------|----------------|---------------|--------------|-----------------|
| `[[` | spatialMap, sampleMap, imgData | spatialMap, experiments | spatialMap, sampleMap, imgData | points, shapes | points, shapes |
| `[` | spatialMap, sampleMap, imgData | spatialMap | spatialMap, sampleMap, imgData, layer lists | points, shapes, spatialMap | points, shapes, spatialMap |
| `nrow` | spatialMap, sampleMap, imgData | spatialMap | spatialMap, sampleMap, imgData | points, shapes, spatialMap | points, shapes, spatialMap |
| `colnames` | spatialMap, imgData | spatialMap, experiments | spatialMap, imgData | points, shapes, spatialMap | points, shapes, spatialMap |
| `names` | — | experiments | layer lists, experiments | layer lists | layer lists |
| `length` | — | — | layer lists | — | — |
| `unique` | — | — | — | spatialMap | spatialMap |
| `split` | — | — | — | spatialMap | spatialMap |
| `st_intersects` (sf) | — | — | — | shapes | points, shapes |
| `st_as_sfc` (sf) | — | — | — | shapes | points, shapes |
| `st_bbox` (sf) | — | — | — | shapes | — |
| `rownames` | colData | — | — | — | — |
| `dim` | — | — | — | — | — |

*Note: `dim` is used in `[` when `drop = TRUE` for experiments. `rownames` is used on `colData` from MultiAssayExperiment.*

**Instance ID type consistency (subsetByBoundingBox, subsetByPolygon):** Spatial filtering matches `instance_id` values between points/shapes elements and `spatialMap` using `%in%`. The package does not coerce types. Ensure `instance_id` in your points, shapes, and `spatialMap` have matching types (e.g., all character or all integer) so the comparison behaves as expected.

## Generics used by subsetByColData

**spatialMap, sampleMap, imgData** (table-like), **colData** (from MultiAssayExperiment):

| Generic | Element | Call pattern | Purpose |
|---------|---------|--------------|---------|
| `nrow` | spatialMap, sampleMap, imgData | `nrow(x) == 0` | Detect empty |
| `colnames` | spatialMap, imgData | `cols %in% colnames(x)` | Check column presence |
| `[[` | spatialMap, sampleMap, imgData | `x[["assay"]]`, `x[["colname"]]`, `x[["sample_id"]]` | Extract columns |
| `[` | spatialMap, sampleMap, imgData | `x[keys %in% other_keys, , drop = FALSE]` | Subset rows |
| `rownames` | colData | `rownames(colData(x))` | Primary IDs for imgData filter |

## Generics used by subsetByColumn

**spatialMap** (table-like), **experiments** (list of assays):

| Generic | Element | Call pattern | Purpose |
|---------|---------|--------------|---------|
| `nrow` | spatialMap | `nrow(spatialMap) == 0` | Detect empty |
| `colnames` | spatialMap, experiments | `colnames(spatialMap)`, `colnames(experiments[[nm]])` | Column/assay names |
| `[[` | spatialMap, experiments | `spatialMap[["assay"]]`, `experiments[[nm]]` | Extract columns, assays |
| `[` | spatialMap | `spatialMap[keys %in% kept, , drop = FALSE]` | Subset rows |
| `names` | experiments | `names(experiments)` | Assay names |

## Generics used by subsetByAssay

**spatialMap, sampleMap, imgData** (table-like), **layer lists** (points, shapes, images, labels):

| Generic | Element | Call pattern | Purpose |
|---------|---------|--------------|---------|
| `nrow` | spatialMap, sampleMap, imgData | `nrow(x) == 0` | Detect empty |
| `colnames` | spatialMap, imgData | `"assay" %in% colnames(x)` | Check columns |
| `[[` | spatialMap, sampleMap | `x[["assay"]]`, `x[["primary"]]` | Extract columns |
| `[` | spatialMap, sampleMap, imgData | `x[x[["assay"]] %in% assays, , drop = FALSE]` | Subset rows |
| `names` | layer lists, experiments | `names(points)`, `names(experiments)` | Layer/assay names |
| `length` | layer lists | `length(names(elements)) == 0` | Check empty |
| `[` | layer lists | `elements[keep]` | Subset layers by name |

## Generics used by subsetByBoundingBox

**Points** (bbox filter):

| Generic | Call pattern | Purpose |
|---------|--------------|---------|
| `[[` | `element[[x_col]]`, `element[[y_col]]` | Extract x, y columns for comparison |
| `colnames` | `x_col %in% colnames(element)` | Check column presence |
| `nrow` | `nrow(element) == 0` | Detect empty |
| `[` | `element[idx, , drop = FALSE]` | Subset rows by integer indices |
| `[[` | `element[["instance_id"]][idx]` | Get instance IDs of retained rows |

The comparison `(xv >= xmin & xv <= xmax) & (yv >= ymin & yv <= ymax)` uses the
vectors returned by `[[`; the result feeds `which()`. Your `[[` method must
return something that supports `>=`, `<=`, and `&` with numeric scalars.

**Shapes** (bbox filter):

| Generic | Call pattern | Purpose |
|---------|--------------|---------|
| `[[` | `element[["geometry"]]` | Extract geometry column |
| `colnames` | `"geometry" %in% colnames(element)` | Check geometry column |
| `nrow` | `nrow(element) == 0` | Detect empty |
| `st_bbox` (sf) | `st_bbox(c(xmin, ymin, xmax, ymax))` | Create bbox from numeric; `st_bbox.numeric` |
| `st_as_sfc` (sf) | `st_as_sfc(bbox)` | Convert bbox to polygon; `st_as_sfc.bbox` |
| `st_intersects` (sf) | `st_intersects(geom, box_sfc)` | Spatial predicate; `st_intersects.sfc` |
| `[` | `element[idx, , drop = FALSE]` | Subset rows |
| `[[` | `element[["instance_id"]][idx]` | Get instance IDs |

The geometry column must be `sfc` (or dispatch to `st_intersects` for your
type). `st_bbox` and `st_as_sfc` are called on a numeric vector and a bbox,
not on your element.

## Generics used by subsetByPolygon

**Points** (polygon filter):

| Generic | Call pattern | Purpose |
|---------|--------------|---------|
| `[[` | `element[[x_col]]`, `element[[y_col]]` | Extract coordinates |
| `colnames` | `x_col %in% colnames(element)` | Check column presence |
| `nrow` | `nrow(element) == 0` | Detect empty |
| `st_as_sfc` (sf) | `st_as_sfc(wkt_strings)` | Build points from WKT; `st_as_sfc.character` |
| `st_intersects` (sf) | `st_intersects(pts_sfc, polygon)` | Point-in-polygon; `st_intersects.sfc` |
| `[` | `element[idx, , drop = FALSE]` | Subset rows |
| `[[` | `element[["instance_id"]][idx]` | Get instance IDs |

Points are converted to WKT via `paste0("POINT(", x, " ", y, ")")`, then
`st_as_sfc.character` builds an `sfc`. Your `[[` must return values that paste
into valid numeric WKT.

**Shapes** (polygon filter):

| Generic | Call pattern | Purpose |
|---------|--------------|---------|
| `[[` | `element[["geometry"]]` | Extract geometry column |
| `colnames` | `"geometry" %in% colnames(element)` | Check geometry column |
| `nrow` | `nrow(element) == 0` | Detect empty |
| `st_intersects` (sf) | `st_intersects(geom, polygon)` | Shape–polygon intersection; `st_intersects.sfc` |
| `[` | `element[idx, , drop = FALSE]` | Subset rows |
| `[[` | `element[["instance_id"]][idx]` | Get instance IDs |

Again, the geometry column is passed to `st_intersects`. Implement
`st_intersects.your_class` if your geometry is not `sfc`, or ensure the column
coerces to `sfc`.

## Summary: what to implement

For **points and shapes** elements (extensible data in MASE):

| Generic | Points (bbox) | Points (polygon) | Shapes (bbox) | Shapes (polygon) |
|---------|---------------|------------------|---------------|------------------|
| `[[` | ✓ | ✓ | ✓ | ✓ |
| `[` | ✓ | ✓ | ✓ | ✓ |
| `colnames` | ✓ | ✓ | ✓ | ✓ |
| `nrow` | ✓ | ✓ | ✓ | ✓ |
| `st_intersects` | — | — | ✓ | ✓ |
| `st_as_sfc` | — | (for WKT) | — | — |
| `st_bbox` | — | — | (for bbox) | — |

For **points**, `[[` must return vectors compatible with numeric comparison
(bbox) or pasteable to WKT (polygon). For **shapes**, `[[` must return geometry
that works with `st_intersects`—either `sfc` or a type for which you add
`st_intersects.your_class`.

## Validity constraints

`PointsLayerList` and `ShapesLayerList` validity require
`is(element, "DataFrame")`. If your class extends `DataFrame`, it will pass.

# Summary

| Use case | Operation |
|----------|-----------|
| Filter by specimen | `subsetByColData` or `x[, j]` with character `j` |
| Filter by assay columns | `subsetByColumn` or `x[, j]` with list `j` |
| Filter by assay | `subsetByAssay` or `x[, , k]` |
| Filter by spatial ROI (rectangle) | `subsetByBoundingBox` |
| Filter by spatial ROI (polygon) | `subsetByPolygon` |
| Compose row/col/assay | `x[i, j, k, drop]` |

For new data representations (points or shapes), implement `[[`, `[`, `colnames`,
and `nrow` for your class. For shapes with non-sfc geometry, add `st_intersects`
(sf) methods so the spatial predicate works. All subsetting uses only S3/S4
generics; see the overview table above for the complete list.

# Session info

```{r sessionInfo}
sessionInfo()
```
